\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-31 paper.tex
Tweaks from Yan

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form, the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages was formalised as a
graph by Griffiths and colleagues~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inference from a set of sampled genome sequences.
The distinction between stochastic process
and data structure is not clear cut, however,
and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Another concept of an ARG that we must be careful to distinguish,
is that of the \emph{actual} ARG underlying a sample of genomes
and ultimately an entire species. This ARG is the true history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Just as this true, ideal ARG unquestionably exists,
it can also never be entirely known,
% NOTE: preempting some discussion here:
% Mutation rate is (a) not high enough in existing species to
% resolve, (b) probably couldn't be if organisms are to survive,
% and (c) such a high mutation rate would probably even obscure
% the more ancient past.
because of a fundamental lack of mutational information.

The scale of modern datasets makes the
relationship between the true ARG and
population genetic models important to articulate.
The coalescent with recombination (henceforth, ``coalescent'')
is a mathematical model that defines a distribution of ARGs
for a given set of parameters.
It has proven to be a useful and robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
but many modern datasets have properties that grossly
violate its assumptions.
One key assumption is that
sample size $n$ is much less than the effective population size, $N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{turnbull2018hundred, bycroft2018genome,
karczewski2020mutational,tanjo2021practical,
halldorsson2022sequences},
and so sample size is an order of magnitude \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from this
assumption, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
A model assuming a single $N_e$ would be a
drastic over-simplification in these cases of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the complexities
of family structure in these
datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
Another core assumption of the coalescent model is that the genome
(or at least the region under study) is short enough that the number of extant
% TODO do we need to explain this a bit more?
lineages remains much smaller than $N_e$ at all times in the past.
High-quality whole genome assemblies are now available
for many species % Can we say something more specific here?
and projects are under way to obtain
them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth}.
Vast quantities of whole genome sequence data across a broad range
of species exists today.

The true ARG underlying a modern population-scale dataset
is therefore not a sample from the coalescent, or any other
population genetics model.

% Mathematical models are insufficient to describe the complexities
% of biological reality when sampling is at population scale,
% and the true ARG can never be known in its entirity.

A reasonable goal to set, then, is to estimate \emph{aspects}
of the true ARG and to capture uncertainty where it inevitably exists.
This is the approach taken by recent inference
methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
by neither fully imposing the constraints of a mathematical
model or trying estimate the true ARG is its compete detail, they
have made substantial progress. However, our vocabularly
and mathematical formalisms, rooted in both the ideal and
mathematical concepts of an ARG, are insufficient to capture the
richness of these new structures.

% This ideal ARG is also not a sample from the coalescent
% or any other population genetics model: by necessity, and design,
% such mathematical constructs are caricatures of biological reality,
% allowing us to draw general conclusions about population processes
% without the burden of unnecessary detail.
% The scale of present-day datasets breaks fundamental assumptions
% made by such models, and we must therefore be careful to articulate
% our goals in inference.

% pedantically, ARG is not really a "word"
The term ``ARG'' is now frequently used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
as a data structure
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage is a useful generalisation,
but is not compatible with the available formal
data structure definitions
derived from the original Griffiths approach.
This mismatch between usage and definition has led to significant
problems, which if not addressed
may impede progress in applying ARGs to understand the biological world.

@@

% A data structure is a means of encoding information, and
% a precise description of the structure allows us to mathematically
% reason about its properties.

% We must also distinguish between a specific realisation
% the \emph{actual} ARG underlying a sample of genomes

@@

% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined here is
in many ways simply a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to describe the
differential inheritance of genetic material between genomes.
% % This is definitely needed for the Nicks, who are always thinking
% % about the deeper processes and are quite happy being loosey-goosey
% % with representation details
It is important to note that here, and throughout,
we are not questioning the form of the
actual ancestral processes that occur in nature, but rather how we
\emph{represent} the outcomes of such processes in a practical manner.
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
These practical details,
as demonstrated in later sections, have important
consequences not only for how methods exchange information about
simulated and inferred ARGs, but more fundamentally in
how we set our goals for inference and evaluate the success of results.

@@

% Fully decoupling the general concept of an ARG from the coalescent
% with recombination (henceforth, ``coalescent'')
% is an important step.
% While the coalescent has proven to be a useful and
% robust
% model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
% many modern datasets have properties that grossly
% violate its assumptions.
% One key assumption is that
% sample size $n$ is much less than the effective population size, $N_e$.
% Several human datasets now consist of hundreds of thousands of
% genomes~\citep{turnbull2018hundred, bycroft2018genome,
% karczewski2020mutational,tanjo2021practical,
% halldorsson2022sequences},
% and so sample size is an order of magnitude \emph{larger} than the
% usually assumed $N_e$ values.
% Agricultural datasets are an even more extreme departure from this
% assumption, with hundreds of thousands of samples embedded in
% multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
% and effective population sizes of 100 and even
% less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
% A model assuming a single $N_e$ would be a
% drastic over-simplification of course, but
% even if sufficiently complex demographic models~\citep{gower2022demes}
% encompassing hundreds of populations, explosive growth rates and myriad
% interconnections of migration, were somehow estimated and provided as input,
% ARGs sampled from the coalescent cannot capture the complexities
% of family structure in these
% datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
% Another core assumption of the coalescent model is that the genome
% (or at least the region under study) is short enough that the number of extant
% lineages remains much smaller than $N_e$ at all times.
% High-quality whole genome assemblies are now available
% for many species % Can we say something more concrete here?
% and projects are under way to obtain
% them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
% and so we can expect inferred ARGs to routinely span whole chromosomes.

>>>>

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form, the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages was formalised as a
graph by Griffiths and colleagues~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inference from a set of sampled genome sequences.
The distinction between stochastic process
and data structure is not clear cut, however,
and subfields use the term
differently (see Appendix~\ref{sec-arg-history}).
% pedantically, ARG is not really a "word"
The term ``ARG'' is now frequently used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
as a data structure
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage is a useful generalisation,
but is not compatible with the available formal definitions
derived from the original Griffiths approach.
This mismatch between usage and definition has led to significant
problems, which if not addressed
may impede progress in applying ARGs to understand the biological world.

@@

@@

% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined here is
in many ways simply a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to describe the
differential inheritance of genetic material between genomes.
% This is definitely needed for the Nicks, who are always thinking
% about the deeper processes and are quite happy being loosey-goosey
% with representation details
It is important to note that here, and throughout,
we are not questioning the form of the
actual ancestral processes that occur in nature, but rather how we
\emph{represent} the outcomes of such processes in a practical manner.
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
These practical details,
as demonstrated in later sections, have important
consequences not only for how methods exchange information about
simulated and inferred ARGs, but more fundamentally in
how we set our goals for inference and evaluate the success of results.

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination (henceforth, ``coalescent'')
is an important step.
While the coalescent has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
One key assumption is that
sample size $n$ is much less than the effective population size, $N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{turnbull2018hundred, bycroft2018genome,
karczewski2020mutational,tanjo2021practical,
halldorsson2022sequences},
and so sample size is an order of magnitude \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from this
assumption, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
A model assuming a single $N_e$ would be a
drastic over-simplification of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the complexities
of family structure in these
datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
Another core assumption of the coalescent model is that the genome
(or at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times.
High-quality whole genome assemblies are now available
for many species % Can we say something more concrete here?
and projects are under way to obtain
them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
and so we can expect inferred ARGs to routinely span whole chromosomes.
